An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-04-25 EconGrowth.Rnw
Additional editing changes.

\noindent where $y \equiv Y/Y_\mathrm{0}$, $k \equiv K/K_\mathrm{0}$, and $l \equiv L/L_\mathrm{0}$; with $Y$ represented by GDP, $K$ expressed in currency units, $L$ expressed in workers/year or work-hours/year, time ($t$) measured in years, e being the base of the natural logarithm, and subscripts ``0'' indicating values at an initial year. $A$ is a technological progress term, with $\lambda$ indicating the pace of technological progress (in units of year$^{-1}$). Neoclassical economic growth theory makes no attempt to understand the mechanisms by which technological progress ($A$) affects economic growth, nor does it attempt to describe how technological progress ($A$) might vary from one economy to the next. Mathematically, $A$ is a fitting parameter that aligns the predictions of Equation \ref{eq:CD} with historical data. Typically, the constraint $\alpha + \beta = 1$ is assumed, indicating constant returns to scale.

@@

This section describes the three coordinates of analysis and reviews studies in the literature related to each.

@@

%------------------------------
\subsubsection{Energy-augmented Cobb-Douglas Production Function (CDe)}
%------------------------------

@@

\citet{Ayres:2008cl} uses Equation \ref{eq:CDe} with exergy ($x$) and useful work ($u$) energy quantifications to describe historic Japanese and American economies from 1900-2000. \citet{Froling:2009vo} uses an energy-augmented Cobb-Douglas production function with thermal energy quantification ($q$) to describe output for the global economy from 1800-1970. Fr\"{o}ling's model endogenizes labor with respect to fertility and mortality rates and separates thermal energy into two types: biomass energy and coal energy. \citet{Serrenho:2010vq} use the log transform of the energy-augmented Cobb-Douglas function with thermal energy quantification ($q$) to describe output for the EU-15 countries. Serrenho divides thermal energy into renewables, non-renewables, wastes, and imports.

@@

%------------------------------
\subsubsection{Energy-augmented CES Production Function (CESe)}
%------------------------------

@@

%------------------------------
\subsubsection{LINEX Production Function}
%------------------------------

@@

%++++++++++++++++++++++++++++++
\subsection{Quantification of Energy}
%++++++++++++++++++++++++++++++

The literature shows three quantifications of energy ($e$) used in energy-augmented production functions: thermal energy ($q$), exergy ($x$), and useful work ($u$). Thermal energy ($q$) is the heat content of an energy carrier. Exergy ($x$) is the maximum useful work that can be extracted from an energy carrier by a process that brings the system into equilibrium with a heat reservoir (typically the surrounding environment). Useful work ($u$) includes the direct energy that goes into an economy, including work from machines, human muscles, and animals. Most energy economists have used thermal energy ($q$) for the energy factor of production ($e$) in energy-augmented production functions, including \citet{Stern:2012ey}, \citet{Froling:2009vo}, \citet{Serrenho:2010vq}, and \citet{Kummel:1980wx,Kummel:1982vy,Kummel:1985vz,Kummel:1989tf,Kummel:2002tx,Kummel:2010vz}. \citet{Warr:2010ga} introduced the idea of using exergy ($x$) and useful work ($u$) in a production function to account for efficiency gains in energy production and consumption over the last century.

We know of no statistically-rigorous comparisons of the predictive ability of each type of energy quantification in the various types of production functions discussed in Section \ref{sec:Prod_Func_Forms}. We analyze all three quantifications of energy ($q$, $x$, and $u$) for developed economies and thermal energy ($q$) and exergy ($x$) for all other economies.

%++++++++++++++++++++++++++++++
\subsection{Economy Types}
%++++++++++++++++++++++++++++++

@@

\begin{table} \caption{Example economies and years of evaluation.} \label{tab:Economy_Examples} 
  \begin{center}
    \begin{tabular}{r c c l l} 
      \multirow{3}{*}{Developed} & \includegraphics[height=0.11in]{flag/flag_US.png} & US & USA & \multirow{3}{*}{1980-2011}\\
                                 & \includegraphics[height=0.11in]{flag/flag_UK.png} & UK & United Kingdom & \\
                                 & \includegraphics[height=0.11in]{flag/flag_JP.png} & JP & Japan          & \\
      \hline
      \multirow{2}{*}{BRICS}     & \includegraphics[height=0.11in]{flag/flag_CN.png} & CN & China & \multirow{2}{*}{1990-2011}        \\
                                 & \includegraphics[height=0.11in]{flag/flag_ZA.png} & ZA & South Africa & \\
      \hline
      \multirow{2}{*}{Petrol}    & \includegraphics[height=0.11in]{flag/flag_SA.png} & SA & Saudi Arabia & \multirow{2}{*}{1990-2011} \\
                                 & \includegraphics[height=0.11in]{flag/flag_IR.png} & IR & Iran         &  \\
      \hline
      \multirow{2}{*}{LDC}       & \includegraphics[height=0.11in]{flag/flag_TZ.png} & TZ & Tanzania & \multirow{2}{*}{1990-2011}     \\
                                 & \includegraphics[height=0.11in]{flag/flag_ZM.png} & ZM & Zambia      & \\
    \end{tabular}
  \end{center}
\end{table}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Models and methodology}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Data sources and data collection}

%++++++++++++++++++++++++++++++
\subsection{Output}
%++++++++++++++++++++++++++++++

We take real gross domestic product to be an appropriate, if imperfect proxy for output. Real gross domestic product data in real 2005 U.S. dollars for the United States, the United Kingdom, and Japan for 1980-2011 is from the World Bank's Global Economic Prospects (GEP) database (WORLD BANK, 2012). Real GDP data in real 2005 U.S. dollars for South Africa, China, Iran, Saudi Arabia, Tanzania, and Zambia for 1991-2011is from the World Bank's Global Economic Prospects database (WORLD BANK, 2012).

%++++++++++++++++++++++++++++++
\subsection{Capital stock}
%++++++++++++++++++++++++++++++

@@

%++++++++++++++++++++++++++++++
\subsection{Labor and employment}
%++++++++++++++++++++++++++++++

@@

%++++++++++++++++++++++++++++++
\subsection{Thermal energy, exergy, and useful work}
%++++++++++++++++++++++++++++++

@@

%++++++++++++++++++++++++++++++
\subsubsection{Thermal energy}
%++++++++++++++++++++++++++++++

@@

%++++++++++++++++++++++++++++++
\subsubsection{Exergy}
%++++++++++++++++++++++++++++++

@@

%------------------------------
\subsubsection{Useful work}
%------------------------------

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Historical Data}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Figure \ref{fig:Historical_Lattice_Graph} shows historical data, including GDP ($y$), capital stock ($k$), labor ($l$), and the various energy quantifications ($q$, $x$, and $u$) for 1980-2011 (developed economies) and 1991-2011 (other economies). Factors of production track each other closely, and large covariances among these variables can be seen in Tables \ref{tab:Covariance_US} through \ref{tab:Covariance_ZM}.

>>>>

\noindent where $y \equiv Y/Y_\mathrm{0}$, $k \equiv K/K_\mathrm{0}$, and $l \equiv L/L_\mathrm{0}$; with $Y$ represented by GDP, $K$ expressed in currency units, $L$ expressed in workers/year or work-hours/year, time ($t$) measured in years, e being the base of the natural logarithm, and subscripts ``0'' indicating values at an initial year.\footnote{Indexed quantities are represented by lower-case symbols ($y$, $k$, $l$, $e$, $q$, $x$, and $u$) and dimensional quantities are represented by upper-case symbols ($Y$, $K$, $L$, $E$, $Q$, $X$, and $U$).} $A$ is a technological progress term, with $\lambda$ indicating the pace of technological progress (in units of year$^{-1}$). Neoclassical economic growth theory makes no attempt to understand the mechanisms by which technological progress ($A$) affects economic growth, nor does it attempt to describe how technological progress ($A$) might vary from one economy to the next. Mathematically, $A$ is a fitting parameter that aligns the predictions of Equation \ref{eq:CD} with historical data. Typically, the constraint $\alpha + \beta = 1$ is assumed, indicating constant returns to scale.

@@

This section describes the three coordinates of analysis and briefly reviews literature related to each.

@@

%------------------------------
\subsubsection{Energy-augmented Cobb-Douglas Production Function (CDe)} \label{sec:CDe}
%------------------------------

@@

\citet{Ayres:2008cl} uses Equation \ref{eq:CDe} with exergy ($x$) and useful work ($u$) quantifications for $e$ to describe historic Japanese and American economies from 1900-2000. \citet{Froling:2009vo} uses an energy-augmented Cobb-Douglas production function with thermal energy quantification ($q$) to describe output for the global economy from 1800-1970. Fr\"{o}ling's model endogenizes labor with respect to fertility and mortality rates and separates thermal energy into two types: biomass energy and coal energy. \citet{Serrenho:2010vq} use the log transform of the energy-augmented Cobb-Douglas function with thermal energy quantification ($q$) to describe output for the EU-15 countries. Serrenho divides thermal energy into renewables, non-renewables, wastes, and imports.

@@

%------------------------------
\subsubsection{Energy-augmented CES Production Function (CESe)} \label{sec:CES}
%------------------------------

@@

%------------------------------
\subsubsection{LINEX Production Function} \label{sec:LINEX}
%------------------------------

@@

%++++++++++++++++++++++++++++++
\subsection{Quantification of Energy} \label{sec:Quantification_Of_Energy}
%++++++++++++++++++++++++++++++

The literature shows three quantifications of energy ($E$) used in energy-augmented production functions: thermal energy ($Q$), exergy ($X$), and useful work ($U$). Thermal energy ($Q$) is the heat content of an energy carrier. Exergy ($X$) is the maximum useful work that can be extracted from an energy carrier by a process that brings the system into equilibrium with a heat reservoir (typically the surrounding environment). Useful work ($U$) includes the direct energy that goes into an economy, including work from machines, human muscles, and animals. Most energy economists have used thermal energy ($Q$) for the energy factor of production ($E$) in energy-augmented production functions, including \citet{Stern:2012ey}, \citet{Froling:2009vo}, \citet{Serrenho:2010vq}, and \citet{Kummel:1980wx,Kummel:1982vy,Kummel:1985vz,Kummel:1989tf,Kummel:2002tx,Kummel:2010vz}. \citet{Warr:2010ga} introduced the idea of using exergy ($X$) and useful work ($U$) in a production function to account for efficiency gains in energy production and consumption over the last century.

We know of no statistically-rigorous comparisons of the predictive ability of each type of energy quantification in the various types of production functions discussed in Section \ref{sec:Prod_Func_Forms}. We analyze all three quantifications of energy ($Q$, $X$, and $U$) for developed economies and thermal energy ($Q$) and exergy ($X$) for all other economies.

%++++++++++++++++++++++++++++++
\subsection{Economy Types} \label{sec:Economy_Types}
%++++++++++++++++++++++++++++++

@@

\begin{table} \caption{Example economies and years of evaluation.} \label{tab:Economy_Examples} 
  \begin{center}
    \begin{tabular}{r c c l l} 
      \multirow{3}{*}{Developed} & \includegraphics[height=0.11in]{flag/flag_US.png} & US & USA & \multirow{3}{*}{1980-2011}\\
                                 & \includegraphics[height=0.11in]{flag/flag_UK.png} & UK & United Kingdom & \\
                                 & \includegraphics[height=0.11in]{flag/flag_JP.png} & JP & Japan          & \\
      \hline
      \multirow{2}{*}{BRICS}     & \includegraphics[height=0.11in]{flag/flag_CN.png} & CN & China & \multirow{2}{*}{1991-2011}        \\
                                 & \includegraphics[height=0.11in]{flag/flag_ZA.png} & ZA & South Africa & \\
      \hline
      \multirow{2}{*}{Petrol}    & \includegraphics[height=0.11in]{flag/flag_SA.png} & SA & Saudi Arabia & \multirow{2}{*}{1991-2011} \\
                                 & \includegraphics[height=0.11in]{flag/flag_IR.png} & IR & Iran         &  \\
      \hline
      \multirow{2}{*}{LDC}       & \includegraphics[height=0.11in]{flag/flag_TZ.png} & TZ & Tanzania & \multirow{2}{*}{1991-2011}     \\
                                 & \includegraphics[height=0.11in]{flag/flag_ZM.png} & ZM & Zambia      & \\
    \end{tabular}
  \end{center}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data Sources} \label{sec:Data_Sources}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To assess energy-augmented models of economic growth along the coordinates of analysis outlined in Section \ref{sec:Coordinates}, historical data for economic output ($Y$), capital stock ($K$), labor ($L$), thermal energy ($Q$), exergy ($X$), and useful work ($U$) is required. Subsections below describe sources for and calculations to obtain historical data. A final subsection presents the data in graphical form.

@@

%++++++++++++++++++++++++++++++
\subsection{Output ($Y$)} \label{sec:Output_Data}
%++++++++++++++++++++++++++++++

We take real gross domestic product to be an appropriate, if imperfect, proxy for output ($Y$). Gross domestic product data in real 2005 U.S. dollars for the United States, the United Kingdom, and Japan for 1980-2011 is from the World Bank's Global Economic Prospects (GEP) database (WORLD BANK, 2012). GDP data in real 2005 U.S. dollars for South Africa, China, Iran, Saudi Arabia, Tanzania, and Zambia for 1991-2011 is from the World Bank's Global Economic Prospects database (WORLD BANK, 2012).

%++++++++++++++++++++++++++++++
\subsection{Capital stock ($K$)} \label{sec:Capital_Stock_Data}
%++++++++++++++++++++++++++++++

@@

%++++++++++++++++++++++++++++++
\subsection{Labor and employment ($L$)} \label{sec:Labor_Data}
%++++++++++++++++++++++++++++++

@@

%++++++++++++++++++++++++++++++
\subsection{Thermal energy, exergy, and useful work ($E$)} \label{sec:Energy_Data}
%++++++++++++++++++++++++++++++

@@

%------------------------------
\subsubsection{Thermal energy ($Q$)} \label{sec:Q_Data}
%------------------------------

@@

%------------------------------
\subsubsection{Exergy ($X$)} \label{sec:X_Data}
%------------------------------

@@

%------------------------------
\subsubsection{Useful work ($U$)} \label{sec:U_Data}
%------------------------------

@@

%++++++++++++++++++++++++++++++
\subsection{Historical Data} \label{sec:Historical_Data}
%++++++++++++++++++++++++++++++

Figure \ref{fig:Historical_Lattice_Graph} presents historical data, including indexed values for output ($y$), capital stock ($k$), labor ($l$), and the various energy quantifications ($q$, $x$, and $u$) for 1980-2011 (developed economies) and 1991-2011 (other economies). Factors of production track each other closely, and large covariances among these variables can be seen in Tables \ref{tab:Covariance_US} through \ref{tab:Covariance_ZM}.
